Exploitation Techniques and Mitigations

### 2016-01-20 docs/paper.tex
Paper: Add section about DEP

Lets start off by examining the stack containing a buffer \texttt{c} as local
variable, see \cref{fig:stack_frame}. Right now the buffer holds the string
\texttt{"hello"} followed by a terminator. Since it has been allocated
to hold a maximum of \SI{12}{\byte} this fits. If data is written to the buffer
larger than \SI{12}{\byte} the following variable (or parameter) \texttt{bar}
will be overwritten, followed by the saved frame pointer and the return
address. If even more data is supplied the following stack frame will be
overwritten in the same manner.

If an attacker can provide the data written to the buffer and no (or wrong)
bound checking is done, he is able to inject arbitrary (malicious) code into
the stack frame. This could be, for instance, be used to overwrite a flag
indication whether an authentication has been performed successfully or not.
But since this is pretty forward lets go beyond that and see what happens when
changing the return address.

@@

As shown in \cref{src:buffer_overflow} we have a buffer suited for
\SI{20}{\byte} but without any bound checking. If the provided input is longer,
we will be able to overwrite the return address. Lets have a look at the
resulting binary utilizing \texttt{objdump}.

Looking at lines 13 and 23 we can infer that the buffer will start
\SI{28}{\byte} (\texttt{0x1c}) before the base pointer. Hence we have to supply
\SI{32}{\byte} (28 + 4) of arbitrary data followed by the address where we want
to jump to. Lets jump into the function \texttt{mordor} located at
\texttt{0x804849b}, keep in mind that the byte order needs to be swapped.

@@

While this is neat and can certainly be useful to an adversary, stack smashing
also enables us to inject arbitrary code into a program. Contrary to the
previous section the target machine will execute code provided by the attacker.
This can be achieved by bending the return address into the buffer used for the
exploit. Provided instructions will be executed upon return. Shell code is a
piece of (binary) code which opens up a shell that reads and executes commands
from an attacker. This example is taken from Dhaval Kapil's
blog\footnote{\url{https://dhavalkapil.com/blogs/Shellcode-Injection/} on
December 2015} there is also a section about this in \textit{Hacking: The Art
of Exploitation}~\cite[pp.~281]{art_of_exploitation}.

@@

Lets first calculate the distance between the start of the buffer and the
return address. The return address will be located \SI{4}{\byte} after the
saved base pointer location.

@@

>>>>

Let's start off by examining the stack containing a buffer \texttt{c} as local
variable, see \cref{fig:stack_frame}. Right now the buffer holds the string
\texttt{"hello"} followed by a terminator. Since it has been allocated
to hold a maximum of \SI{12}{\byte} this fits. If data is written to the buffer
larger than \SI{12}{\byte} the following variable (or parameter) \texttt{bar}
will be overwritten, followed by the saved frame pointer and the return
address. If even more data is supplied the following stack frame will be
overwritten in the same manner.

If an attacker can provide the data written to the buffer and no (or wrong)
bound checking is done, he is able to inject arbitrary (malicious) code into
the stack frame. This could be, for instance, be used to overwrite a flag
indication whether an authentication has been performed successfully or not.
But since this is pretty forward let's go beyond that and see what happens when
changing the return address.

@@

As shown in \cref{src:buffer_overflow} we have a buffer suited for
\SI{20}{\byte} but without any bound checking. If the provided input is longer,
we will be able to overwrite the return address. Let's have a look at the
resulting binary utilizing \texttt{objdump}.

Looking at lines 13 and 23 we can infer that the buffer will start
\SI{28}{\byte} (\texttt{0x1c}) before the base pointer. Hence we have to supply
\SI{32}{\byte} (28 + 4) of arbitrary data followed by the address where we want
to jump to. Let's jump into the function \texttt{mordor} located at
\texttt{0x804849b}, keep in mind that the byte order needs to be swapped.

@@

While this is neat and can certainly be useful to an adversary, stack smashing
also enables us to inject arbitrary code into a program. Contrary to the
previous section the target machine will execute code provided by the attacker.
This can be achieved by bending the return address into the buffer used for the
exploit. Provided instructions will be executed upon return. Shell code is a
piece of (binary) code which opens up a shell that reads and executes commands
from an attacker. This example is taken from Dhaval Kapil's
blog\footnote{\url{https://dhavalkapil.com/blogs/Shellcode-Injection/} on
December 2015} there is also a section about this in \textit{Hacking: The Art
of Exploitation}~\cite[pp.~281]{art_of_exploitation}. There is also a
comprehensive article \cite{phrack_stack_smash} about Stack Smashing on
\texttt{phrack}\footnote{\url{http://phrack.org/}}.

@@

Let's first calculate the distance between the start of the buffer and the
return address. The return address will be located \SI{4}{\byte} after the
saved base pointer location.

@@

The first mitigation technique discussed, DEP, is also known under the term
\textit{write XOR execute (\texttt{w\^{}x})} and it will prevent us from
executing injected code like we did in the shell code example previously. This
happens by attaching a execute flag to each page (in addition to the read /
write flags). If the instruction pointer points to a page without the execute
flag set, a segmentation fault will be triggered. The only pages flagged for
execution are the ones that belong either to the \texttt{.text} segment or the
used libraries since it will contain the program code. The stack is (of course)
not executable per default. So our shell code example would simply segfault.

On modern platforms this is enforced by hardware, but even CPUs that lack such
hardware support, software-enforced DEP provides limited
protection.~\cite{wiki:dep}

Previously we worked around this by passing \texttt{-z execstack}, hence data
on the stack was executable. This can also be seen in the output of
\texttt{readelf}.

\begin{listing}[h!]
    \codefile{../src/shell_code/vuln_readelf.txt}
    \caption{\texttt{readelf} output with and without \texttt{-z execstack}}
    \label{src:dep_readelf}
\end{listing}

\begin{listing}[h!]
    \codefile{../src/shell_code/run_dep.txt}
    \caption{Running the shell code example without \texttt{-z execstack}}
    \label{src:dep_shell_code}
\end{listing}

\Cref{src:dep_readelf} shows that with \texttt{-z execstack} the stack section
is marked (looking at \texttt{Flg} in lines 8 and 19) with execute
(\texttt{E}). Running the example, see \cref{src:dep_shell_code}, yields the
segfault described in the previous paragraph. The program is terminated upon
receiving the segfault.

DEP makes injection arbitrary code into binary much harder, but we still
control the return address.
